Causal inference is not a statistical problem

### 2023-04-03 manuscript.qmd
update manuscript

---
title: "Causal inference is not a statistical problem"
author: "Lucy D'Agostino McGowan"
format: pdf
execute: 
  echo: false
  message: false
  warning: false
editor_options: 
  chunk_output_type: console
---

@@

Anscombe's quartet is a set of four datasets that have the same statistical properties in terms of summary statistics (means, variances, correlations, and linear regression fits), but exhibit different distributions and relationships when plotted on a graph. The quartet, often used to teach introductory statistics courses, was created to illustrate the importance of visualizing data before drawing conclusions based on statistical analyses alone. Here, we propose a different quartet, where again statistical summaries do not provide insight into the true underlying mechanism, but even visualizations do not solve the issue. Here, an understanding or assumption of the data generating mechanism is required to correctly capture the relationship between the available factors. This proposed quartet is meant to help readers better understand the assumptions underlying causal inference methods, further driving home the point that in order to accurately estimate causal effects we require more information than can be gleaned from statistical tools alone.

@@

We propose the following four data generation mechanisms, summarized by the equations below as well as the directed acyclic graphs displayed in @fig-1. Here, $X$ is presumed to be some exposure of interest, $Y$ an outcome, and $Z$ a known, measured factor. The M-Bias equation includes two additional unmeasured factors, $U_1$ and $U_2$.

@@

$$X \sim N(0, 1)$$

$$Y = X + \varepsilon_y, \textrm{ }\varepsilon_y\sim N(0, 1)$$

$$Z =  0.45X + 0.77 Y + \varepsilon_z$$, $$\varepsilon_z \sim N(0,1)$$

@@

$$Z \sim N(0, 1)$$

$$X = Z + \varepsilon_x,\textrm{ }\varepsilon_x\sim N(0, 1)$$

$$Y =  0.5Z + \varepsilon_y, \textrm{ }\varepsilon_y\sim N(0, 1)$$

@@

$$X \sim N(0, 1)$$

$$Z = X + \varepsilon_z, \textrm{ }\varepsilon_z\sim N(0, 1)$$

$$Y =  Z + \varepsilon_y, \textrm{ }\varepsilon_y\sim N(0, 1)$$

@@

$$U_1 \sim N(0, 1)$$

$$U_2 \sim N(0, 1)$$

$$Z = 8 U_1 + U_2 + \varepsilon_z, \textrm{ }\varepsilon_z\sim N(0, 1)$$

$$X =  U_1 + \varepsilon_x, \textrm{ }\varepsilon_x\sim N(0, 1)$$

$$Y =  X + U_2 + \varepsilon_y, \textrm{ }\varepsilon_y\sim N(0, 1)$$

@@

set.seed(472046)

@@

+---------------------------+----------------------------+-----------------------+
| Data generating mechanism | Correct causal model       | Correct causal effect |
+===========================+============================+=======================+
| \(1\) Collider            | Y \~ X                     | 1                     |
+---------------------------+----------------------------+-----------------------+
| \(2\) Confounder          | Y \~ X + Z                 | 0.5                   |
+---------------------------+----------------------------+-----------------------+
| \(3\) Mediator            | Direct effect: Y \~ X + Z  | Direct effect: 0      |
|                           |                            |                       |
|                           | Total Effect: Y \~ X       | Total effect: 1       |
+---------------------------+----------------------------+-----------------------+
| \(4\) M-Bias              | Y \~ X                     | 1                     |
+---------------------------+----------------------------+-----------------------+

@@

Here we have demonstrated that when presented with an exposure, outcome, and some measured factors, statistics alone, whether summary statistics or data visualizations, are not sufficient to determine the appropriate causal estimate. Additional information about the data generating mechanism is needed in order to draw the correct conclusions.

## Appendix

>>>>

---
title: "Causal inference is not a statistical problem"
author: "Lucy D'Agostino McGowan"
format: pdf
execute: 
  echo: false
  message: false
  warning: false
editor_options: 
  chunk_output_type: console
bibliography: citations.bib
---

@@

Anscombe's quartet is a set of four datasets that have the same statistical properties in terms of summary statistics (means, variances, correlations, and linear regression fits), but exhibit different distributions and relationships when plotted on a graph [@anscombe1973graphs]. The quartet, often used to teach introductory statistics courses, was created to illustrate the importance of visualizing data before drawing conclusions based on statistical analyses alone. Here, we propose a different quartet, where again statistical summaries do not provide insight into the true underlying mechanism, but even visualizations do not solve the issue. In these examples an understanding or assumption of the data generating mechanism is required to correctly capture the relationship between the available factors. This proposed quartet is meant to help practitioners better understand the assumptions underlying causal inference methods, further driving home the point that in order to accurately estimate causal effects we require more information than can be gleaned from statistical tools alone.

@@

We propose the following four data generation mechanisms, summarized by the equations below as well as the directed acyclic graphs displayed in @fig-1. Here, $X$ is presumed to be some continuous exposure of interest, $Y$ a continuous outcome, and $Z$ a known, measured factor. The M-Bias equation includes two additional unmeasured factors, $U_1$ and $U_2$.

@@

$$
\begin{split}
X &\sim N(0, 1)\\
Y &= X + \varepsilon_y, \textrm{ }\varepsilon_y\sim N(0, 1)\\
Z &=  0.45X + 0.77 Y + \varepsilon_z, \textrm{ }\varepsilon_z \sim N(0,1)
\end{split}
$$ {#eq-col}

@@

$$
\begin{split}
Z &\sim N(0, 1)\\
X &= Z + \varepsilon_x,\textrm{ }\varepsilon_x\sim N(0, 1)\\
Y &=  0.5Z + \varepsilon_y, \textrm{ }\varepsilon_y\sim N(0, 1)
\end{split}
$$ {#eq-conf}

@@

$$
\begin{split}
X &\sim N(0, 1)\\
Z &= X + \varepsilon_z, \textrm{ }\varepsilon_z\sim N(0, 1)\\
Y &=  Z + \varepsilon_y, \textrm{ }\varepsilon_y\sim N(0, 1)
\end{split}
$$ {#eq-med}

@@

$$
\begin{split}
U_1 &\sim N(0, 1)\\
U_2 &\sim N(0, 1)\\
Z &= 8 U_1 + U_2 + \varepsilon_z, \textrm{ }\varepsilon_z\sim N(0, 1)\\
X &=  U_1 + \varepsilon_x, \textrm{ }\varepsilon_x\sim N(0, 1)\\
Y &=  X + U_2 + \varepsilon_y, \textrm{ }\varepsilon_y\sim N(0, 1)
\end{split}
$$ {#eq-mbias}

@@

set.seed(229655)

@@

+---------------------------+---------------------------+-----------------------+
| Data generating mechanism | Correct causal model      | Correct causal effect |
+===========================+===========================+=======================+
| \(1\) Collider            | Y \~ X                    | 1                     |
+---------------------------+---------------------------+-----------------------+
| \(2\) Confounder          | Y \~ X + Z                | 0.5                   |
+---------------------------+---------------------------+-----------------------+
| \(3\) Mediator            | Direct effect: Y \~ X + Z | Direct effect: 0      |
|                           |                           |                       |
|                           | Total Effect: Y \~ X      | Total effect: 1       |
+---------------------------+---------------------------+-----------------------+
| \(4\) M-Bias              | Y \~ X                    | 1                     |
+---------------------------+---------------------------+-----------------------+

@@

Here we have demonstrated that when presented with an exposure, outcome, and some measured factors, statistics alone, whether summary statistics or data visualizations, are not sufficient to determine the appropriate causal estimate. Additional information about the data generating mechanism is needed in order to draw the correct conclusions. While knowledge of the data generating process is necessary in order to estimate the correct causal effect in each of the cases presented, there are steps an analyst can take in order to make mistakes such as those presented here less likely. The first is considering the time ordering of the available factors. Fundamental principles of causal inference dictate that the exposure of interest must precede the outcome of interest to plausibly establish a causal relationship. In addition, to account for potential confounding, any covariates adjusted for in the analysis must precede the exposure in time. Including this additional timing information would omit the potential for two of the three misspecified models above (@eq-col the "collider" and @eq-med the "mediator") as the former would demonstrate that the factor $Z$ falls after both the exposure and outcome and the latter would demonstrate that the factor $Z$ falls between the exposure and the outcome in time. Indeed, adjusting for only pre-exposure factors is widely recommended. The only exception is when a known confounder is only measured after the exposure in a particular data analysis, in which case some experts recommend adjusting for it, but even then, caution is advised. [@groenwold2021adjust] In fact many causal inference methodologists would recommend conditioning on *all* measured pre-exposure factors. [@rosenbaumconstructing; @rubin2009should; @rubin1996matching; @rubin2008objective] Including timing information alone (and thus adjusting for all pre-exposure factors) does not preclude one from mistakenly fitting the adjusted model under the fourth data generating mechanism (M-bias), as $Z$ can fall temporally prior to $X$ and $Y$ and still induce bias. It has been argued, however, that this strict M-bias (e.g. where $U_1$ and $U_2$ in @eq-mbias have no relationship with each other and $Z$ has no relationship with $X$ or $Y$ other than via $U_1$ and $U_2$) is very rare in most practical settings. [@liu2012implications; @rubin2009should; @gelman2011causality] Indeed, even theoretical results have demonstrated that bias induced by this data generating mechanism is very sensitive to any deviations from this form. [@ding2015adjust]

We have presented a quartet of examples demonstrating the importance of understanding the data-generating mechanism when attempting to answer causal questions. These data demonstrate that statistical summaries and visualizations alone will not provide insight into the true underlying relationship between the variables, and that an understanding or assumption of the data-generating mechanism is required to correctly capture causal relationships. These examples serve as a reminder of the limitations of statistical tools and the importance of domain-specific knowledge in data analyses.

## References

::: {#refs}
:::

## Appendix {.appendix}
